---
title: "Statistical Modeling Project"
author: "Prateek Sethi"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# STATISTICAL MODELLING PROJECT

## Regression Analysis on the Concrete Compressive Strength Dataset 

#### Load required packages
```{r}
library("readxl")
library(faraway)
library(ggplot2)
library(reshape2)
library(lmtest) 
library(dplyr)
library(ggplot2)
library(randomForest)
library(MASS)
```


#### Load dataset
```{r}
data <- read_excel("Concrete_Data.xls")
```


#### Change column names
```{r}
#changing column names to make them shorter
colnames(data)[colnames(data) == "Cement (component 1)(kg in a m^3 mixture)"] ="cement"
colnames(data)[colnames(data) == "Blast Furnace Slag (component 2)(kg in a m^3 mixture)"] ="blast_furnace_slag"
colnames(data)[colnames(data) == "Fly Ash (component 3)(kg in a m^3 mixture)"] ="fly_ash"
colnames(data)[colnames(data) == "Superplasticizer (component 5)(kg in a m^3 mixture)"] ="superplasticizer"
colnames(data)[colnames(data) == "Water  (component 4)(kg in a m^3 mixture)"] ="water"
colnames(data)[colnames(data) == "Coarse Aggregate  (component 6)(kg in a m^3 mixture)"] ="coarse_agg"
colnames(data)[colnames(data) == "Fine Aggregate (component 7)(kg in a m^3 mixture)"] ="fine_agg"
colnames(data)[colnames(data) == "Age (day)"] ="age"
colnames(data)[colnames(data) == "Concrete compressive strength(MPa, megapascals)"] ="concrete_strength"

```


### Univariate Analysis

#### Data Description
```{r}
#column names
names(data)

#data size
dim(data)  

#variable description
str(data)

#top 5 rows
head(data)

#data summary
summary(data)
```


#### Data Distribution
```{r}
par(mfrow=c(3,3))
hist(data$cement)
hist(data$blast_furnace_slag)
hist(data$fly_ash)
hist(data$water)
hist(data$superplasticizer)
hist(data$coarse_agg)
hist(data$fine_agg)
hist(data$age)
hist(data$concrete_strength)
```


### Multivariate Analysis

#### Correlation Matrix
```{r fig2, fig.height = 30, fig.width = 30}
pairs(data)
```


#### Correlation Heatmap
```{r}
corr_mat <- round(cor(data),2)
melted_corr_mat <- melt(corr_mat)
ggplot(data = melted_corr_mat, aes(x=Var1, y=Var2,
                                   fill=value), size = 10) +
  geom_tile() +
  geom_text(aes(Var2, Var1, label = value),
            color = "white", size = 3)
```


#### Split train test set
```{r}
seed = 21

##train test split
set.seed(seed)
dt = sort(sample(nrow(data), nrow(data)*.80))
train<-data[dt,]
test<-data[-dt,]
```


#### Baseline Model
```{r}
baseline = lm(concrete_strength~.,data=train)
```



#### Variance Inflation Factor
```{r}
vif(baseline)
```


#### Outliers
```{r}
nrow(data[which(abs(rstandard(baseline)) > 2) , ])
```


#### Influential Points
```{r}
indices = cooks.distance(baseline) > 4 / length(cooks.distance(baseline))
nrow(data[which(indices) , ]) 
```


#### Test Model Assumptions
```{r}
plot_residuals <- function(model) {
  plot(fitted(model), resid(model), col = "grey", pch = 20,
       xlab = "Fitted", ylab = "Residuals", main = "Resid plot")
  abline(h = 0, col = "darkorange", lwd = 2)
}

plot_qq <- function(model) {
  qqnorm(resid(model))
  qqline(resid(model), col = "dodgerblue", lwd = 2)
}

check_model_assumptions <- function(model) {
  #check by graphs
  plot_residuals(model)
  #invisible(readline(prompt="Press [enter] to continue"))
  plot_qq(model)
  
  #bptest for equal variance
  print(bptest(model))
  
  #shapiro wilk test for normality
  print(shapiro.test(resid(model)))
}

check_model_assumptions(baseline)
```

### Applying Response Transformation to satisfy model assumptions

#### Using BoxCox graph
```{r}
par(mfrow=c(1,1))
boxcox(concrete_strength~.,data=train)
```
lambda  =  0.8

#### Transform response variable and check model assumptions
```{r}
lambda  =  0.8
transformed_model = lm(((concrete_strength^lambda)-1)/lambda ~., data=train)
summary(transformed_model)

check_model_assumptions(transformed_model)
```
The model assumptions still fail for linearity, normality and equal variance. 

## Prediction Performance

### Variable Addition [Increasing model complexity]

We will now start adding non linear predictor variables in order to capture any nonlinearity that's present in the data.

#### Fit a simple model with all predictors and calculate MSE and PRESS score
```{r}
mlr <- lm(concrete_strength~.,data=train)
summary(mlr)


n = nrow(train)
press_mlr = sqrt(sum((resid(mlr)/(1-hatvalues(mlr)))^2)/n)
press_mlr


calculate_mse_test <- function(model, test) {
  ypred = predict(model, newdata = test)
  resid = ypred - test$concrete_strength
  mse = mean(resid^2)
  return(mse)
}

mse_mlr = calculate_mse_test(mlr, test)
mse_mlr
```


#### Add squared polynomial terms and calculate MSE and PRESS score
```{r}
mlr_squared <- lm(concrete_strength ~ cement + blast_furnace_slag + 
                    fly_ash + water + superplasticizer + coarse_agg + 
                    fine_agg + age + I(cement^2) + I(blast_furnace_slag^2) + 
                    I(fly_ash^2) + I(water^2) + I(superplasticizer^2) + 
                    I(coarse_agg^2) + 
                    I(fine_agg^2) + I(age^2), data = train)
summary(mlr_squared)

press_sq = sqrt(sum((resid(mlr_squared)/(1-hatvalues(mlr_squared)))^2)/n)
press_sq

mse_sq = calculate_mse_test(mlr_squared, test)
mse_sq
```


#### Add cubic polynomial terms and calculate MSE and PRESS score
```{r}
mlr_cubed <- lm(concrete_strength ~ cement + blast_furnace_slag + 
                  fly_ash + water + superplasticizer + coarse_agg + 
                  fine_agg + age + I(cement^2) + I(blast_furnace_slag^2) + 
                  I(fly_ash^2) + I(water^2) + I(superplasticizer^2) + 
                  I(coarse_agg^2) + 
                  I(fine_agg^2) + I(age^2) + I(cement^3) + 
                  I(blast_furnace_slag^3) + 
                  I(fly_ash^3) + I(water^3) + I(superplasticizer^3) + 
                  I(coarse_agg^3) + 
                  I(fine_agg^3) + I(age^3), data = train)
summary(mlr_cubed)

press_cub = sqrt(sum((resid(mlr_cubed)/(1-hatvalues(mlr_cubed)))^2)/n)
press_cub

mse_cub = calculate_mse_test(mlr_cubed, test)
mse_cub
```


#### Add square root terms and calculate MSE and PRESS score
```{r}
mlr_sqrt <- lm(concrete_strength ~ cement + blast_furnace_slag + 
                 fly_ash + water + superplasticizer + coarse_agg + 
                 fine_agg + age + I(cement^2) + I(blast_furnace_slag^2) + 
                 I(fly_ash^2) + I(water^2) + I(superplasticizer^2) + 
                 I(coarse_agg^2) + 
                 I(fine_agg^2) + I(age^2) + I(cement^3) + 
                 I(blast_furnace_slag^3) + 
                 I(fly_ash^3) + I(water^3) + I(superplasticizer^3) + 
                 I(coarse_agg^3) + 
                 I(fine_agg^3) + I(age^3) + I(sqrt(cement)) + 
                 I(sqrt(blast_furnace_slag)) + 
                 I(sqrt(fly_ash)) + I(sqrt(water)) + 
                 I(sqrt(superplasticizer)) + I(sqrt(coarse_agg)) + 
                 I(sqrt(fine_agg)) + I(sqrt(age)), data = train)
summary(mlr_sqrt)

press_sqrt = sqrt(sum((resid(mlr_sqrt)/(1-hatvalues(mlr_sqrt)))^2)/n)
press_sqrt

mse_sqrt = calculate_mse_test(mlr_sqrt, test)
mse_sqrt
```



#### Add logarithmic terms and calculate MSE and PRESS score
```{r}
mlr_log <- lm(concrete_strength ~ cement + blast_furnace_slag + 
                fly_ash + water + superplasticizer + coarse_agg + 
                fine_agg + age + I(cement^2) + I(blast_furnace_slag^2) + 
                I(fly_ash^2) + I(water^2) + I(superplasticizer^2) + 
                I(coarse_agg^2) + 
                I(fine_agg^2) + I(age^2) + I(cement^3) + 
                I(blast_furnace_slag^3) + 
                I(fly_ash^3) + I(water^3) + I(superplasticizer^3) + 
                I(coarse_agg^3) + 
                I(fine_agg^3) + I(age^3) + I(sqrt(cement)) + 
                I(sqrt(blast_furnace_slag)) + 
                I(sqrt(fly_ash)) + I(sqrt(water)) + 
                I(sqrt(superplasticizer))+ I(sqrt(coarse_agg)) + 
                I(sqrt(fine_agg)) + I(sqrt(age)) + I(log(cement+1)) + 
                I(log(blast_furnace_slag+1)) + 
                I(log(fly_ash+1)) + I(log(water+1)) + 
                I(log(superplasticizer+1)) + I(log(coarse_agg + 1)) + 
                I(log(fine_agg+1)) + I(log(age + 1)) , data = train)
summary(mlr_log)

press_log = sqrt(sum((resid(mlr_log)/(1-hatvalues(mlr_log)))^2)/n)
press_log

mse_log = calculate_mse_test(mlr_log, test)
mse_log
```


#### Adding 2nd order interaction terms 
```{r}
mlr_int <- lm(concrete_strength ~ cement + blast_furnace_slag + fly_ash + 
                water + superplasticizer + coarse_agg + fine_agg + age +
                cement*blast_furnace_slag + cement*fly_ash + cement*water +
                cement*superplasticizer + cement*coarse_agg 
                +cement*fine_agg+
                cement*age + blast_furnace_slag*fly_ash + 
                blast_furnace_slag*fly_ash+
                blast_furnace_slag*water + 
                blast_furnace_slag*superplasticizer + 
                blast_furnace_slag*coarse_agg + blast_furnace_slag*fine_agg+
                blast_furnace_slag*age + fly_ash*water + 
                fly_ash*superplasticizer + 
                fly_ash*coarse_agg + fly_ash*fine_agg + fly_ash*age + 
                water*superplasticizer + water*coarse_agg + water*fine_agg+
                water*age + superplasticizer*coarse_agg + 
                superplasticizer*fine_agg+
                superplasticizer*age + coarse_agg*fine_agg + coarse_agg*age+
                fine_agg*age , data = train)
summary(mlr_int)

press_int = sqrt(sum((resid(mlr_int)/(1-hatvalues(mlr_int)))^2)/n)
press_int

mse_int = calculate_mse_test(mlr_int, test)
mse_int
```


#### Add both polynomial and 2nd order interaction terms
```{r}
mlr_allt_int <- lm(concrete_strength ~ cement + blast_furnace_slag + fly_ash
                  + water + superplasticizer + coarse_agg + fine_agg + age +
                  cement*blast_furnace_slag + cement*fly_ash + cement*water+
                  cement*superplasticizer + cement*coarse_agg 
                  +cement*fine_agg +
                  cement*age + blast_furnace_slag*fly_ash + 
                  blast_furnace_slag*fly_ash+
                  blast_furnace_slag*water + 
                  blast_furnace_slag*superplasticizer + 
                  blast_furnace_slag*coarse_agg + 
                  blast_furnace_slag*fine_agg
                  +blast_furnace_slag*age + fly_ash*water + 
                  fly_ash*superplasticizer + 
                  fly_ash*coarse_agg + fly_ash*fine_agg + fly_ash*age + 
                  water*superplasticizer + water*coarse_agg + 
                  water*fine_agg+
                  water*age + superplasticizer*coarse_agg + 
                  superplasticizer*fine_agg+
                  superplasticizer*age + coarse_agg*fine_agg + 
                  coarse_agg*age
                  +fine_agg*age + I(cement^2) + I(blast_furnace_slag^2) + 
                  I(fly_ash^2) + I(water^2) + I(superplasticizer^2) + 
                  I(coarse_agg^2) + I(fine_agg^2) + I(age^2) + I(cement^3) +
                  I(blast_furnace_slag^3) + I(fly_ash^3) + I(water^3) + 
                  I(superplasticizer^3) + I(coarse_agg^3) + I(fine_agg^3) + 
                  I(age^3) + I(sqrt(cement)) + I(sqrt(blast_furnace_slag)) +
                  I(sqrt(fly_ash)) + I(sqrt(water)) + 
                  I(sqrt(superplasticizer)) + I(sqrt(coarse_agg)) + 
                  I(sqrt(fine_agg)) + I(sqrt(age)) + I(log(cement+1)) + 
                  I(log(blast_furnace_slag+1)) + 
                  I(log(fly_ash+1)) + I(log(water+1)) + 
                  I(log(superplasticizer+1)) + I(log(coarse_agg)) + 
                  I(log(fine_agg+1)) + I(log(age)), data = train)

summary(mlr_allt_int)

press_allt_int = sqrt(sum((resid(mlr_allt_int)/(1-hatvalues(mlr_allt_int)))^2)/n)
press_allt_int

mse_all = calculate_mse_test(mlr_allt_int, test)
mse_all
```



#### Plotting model metrics
```{r}
tab <- matrix(c('M1', 'M2', 'M3', 'M4', 'M5', 'M6', 'M7',
                'Baseline', 'Baseline+Squared', 'Baseline+Squared+Cubic', 
                'Baseline+Squared+Cubic+SquareRoot', 
                'Baseline+Squared+Cubic+SquareRoot+Log' , 
                'Baseline+Interaction', 
                'Baseline+Squared+Cubic+SquareRoot+Log+Interaction',
                mlr$rank-1, mlr_squared$rank-1, mlr_cubed$rank-1, 
                mlr_sqrt$rank-1, mlr_log$rank-1,
                mlr_int$rank-1, mlr_allt_int$rank-1, 
                summary(mlr)$r.squared, summary(mlr_squared)$r.squared, 
                summary(mlr_cubed)$r.squared, summary(mlr_sqrt)$r.squared, 
                summary(mlr_log)$r.squared,
                summary(mlr_int)$r.squared, summary(mlr_allt_int)$r.squared,
                summary(mlr)$adj.r.squared, 
                summary(mlr_squared)$adj.r.squared, 
                summary(mlr_cubed)$adj.r.squared, 
                summary(mlr_sqrt)$adj.r.squared, 
                summary(mlr_log)$adj.r.squared,
                summary(mlr_int)$adj.r.squared, 
                summary(mlr_allt_int)$adj.r.squared,
                mse_mlr, mse_sq, mse_cub, mse_sqrt, mse_log, mse_int, 
                mse_all,
                press_mlr, press_sq, press_cub, press_sqrt, press_log, 
                press_int, press_allt_int
                ), ncol=7)
colnames(tab) <- c('Model_Name', 'Model_Description','No_of_predictors','R2',                 'Adj_R2', 'MSE', 'PRESS')
tab <- as.table(tab)

metrics_df = as.data.frame.matrix(tab) 

metrics_df$R2 = as.numeric(as.character(metrics_df$R2))
metrics_df$Adj_R2 = as.numeric(as.character(metrics_df$Adj_R2))
metrics_df$MSE = as.numeric(as.character(metrics_df$MSE))
metrics_df$PRESS = as.numeric(as.character(metrics_df$PRESS))

metrics_df = metrics_df %>% mutate(across(is.numeric, round, digits=2))


ggplot(metrics_df, aes(x=Model_Name, y=as.numeric(R2))) + geom_line(aes(group=1)) + geom_point()

ggplot(metrics_df, aes(x=Model_Name, y=as.numeric(Adj_R2))) + geom_line(aes(group=1)) + geom_point()

ggplot(metrics_df, aes(x=Model_Name, y=as.numeric(MSE))) + geom_line(aes(group=1)) + geom_point()

ggplot(metrics_df, aes(x=Model_Name, y=as.numeric(PRESS))) + geom_line(aes(group=1)) + geom_point()

```




### Variable Reduction 

We will now attempt to reduce the number of variables (model complexity) by penalizing predictors in the full model(both polynomial and interaction) in order to reduce overfitting.

#### Backward AIC Regression
```{r}
fit_back_aic = step(mlr_allt_int, direction = "backward", trace = 0)
summary(fit_back_aic)

press_fit_back_aic = sqrt(sum((resid(fit_back_aic)/(1-hatvalues(fit_back_aic)))^2)/n)
press_fit_back_aic

mse_fbaic = calculate_mse_test(fit_back_aic, test)
mse_fbaic
```


#### Backward BIC Regression
```{r}
n = nrow(train)
fit_back_bic = step(mlr_allt_int, direction = "backward", k=log(n), trace = 0)
summary(fit_back_bic)

press_fit_back_bic = sqrt(sum((resid(fit_back_bic)/(1-hatvalues(fit_back_bic)))^2)/n)
press_fit_back_bic

mse_fbbic = calculate_mse_test(fit_back_bic, test)
mse_fbbic
```


#### Forward AIC Regression
```{r}

fit_null = lm(concrete_strength~1,data=train)
fit_forw_aic = step(fit_null, 
                  scope = concrete_strength ~ cement + blast_furnace_slag +
                  fly_ash 
                  + water + superplasticizer + coarse_agg + fine_agg + age +
                  cement*blast_furnace_slag + cement*fly_ash + cement*water 
                  +
                  cement*superplasticizer + cement*coarse_agg 
                  +cement*fine_agg +
                  cement*age + blast_furnace_slag*fly_ash + 
                  blast_furnace_slag*fly_ash+
                  blast_furnace_slag*water + 
                  blast_furnace_slag*superplasticizer + 
                  blast_furnace_slag*coarse_agg + 
                  blast_furnace_slag*fine_agg
                  +blast_furnace_slag*age + fly_ash*water + 
                  fly_ash*superplasticizer + 
                  fly_ash*coarse_agg + fly_ash*fine_agg + fly_ash*age + 
                  water*superplasticizer + water*coarse_agg + 
                  water*fine_agg+
                  water*age + superplasticizer*coarse_agg + 
                  superplasticizer*fine_agg+
                  superplasticizer*age + coarse_agg*fine_agg + 
                  coarse_agg*age
                  +fine_agg*age + I(cement^2) + I(blast_furnace_slag^2) + 
                  I(fly_ash^2) + I(water^2) + I(superplasticizer^2) + 
                  I(coarse_agg^2) + I(fine_agg^2) + I(age^2) + I(cement^3) + 
                  I(blast_furnace_slag^3) + I(fly_ash^3) + I(water^3) + 
                  I(superplasticizer^3) + I(coarse_agg^3) + I(fine_agg^3) + 
                  I(age^3) + I(sqrt(cement)) + I(sqrt(blast_furnace_slag)) + 
                  I(sqrt(fly_ash)) + I(sqrt(water)) + 
                  I(sqrt(superplasticizer)) + I(sqrt(coarse_agg)) + 
                  I(sqrt(fine_agg)) + I(sqrt(age)) + I(log(cement+1)) + 
                  I(log(blast_furnace_slag+1)) + 
                  I(log(fly_ash+1)) + I(log(water+1)) + 
                  I(log(superplasticizer+1)) + I(log(coarse_agg)) + 
                  I(log(fine_agg+1)) + I(log(age)),   
                  direction = "forward", trace = 0)

summary(fit_forw_aic)

press_fit_forw_aic = sqrt(sum((resid(fit_forw_aic)/(1-hatvalues(fit_forw_aic)))^2)/n)
press_fit_forw_aic

mse_ffaic = calculate_mse_test(fit_forw_aic, test)
mse_ffaic
```


#### Forward BIC Regression
```{r}
fit_null = lm(concrete_strength~1,data=train)
fit_forw_bic = step(fit_null, 
                  scope = concrete_strength ~ cement + blast_furnace_slag +
                  fly_ash 
                  + water + superplasticizer + coarse_agg + fine_agg + age +
                  cement*blast_furnace_slag + cement*fly_ash + cement*water +
                  cement*superplasticizer + cement*coarse_agg 
                  +cement*fine_agg +
                  cement*age + blast_furnace_slag*fly_ash + 
                  blast_furnace_slag*fly_ash+
                  blast_furnace_slag*water + 
                  blast_furnace_slag*superplasticizer + 
                  blast_furnace_slag*coarse_agg + blast_furnace_slag*fine_agg
                  +blast_furnace_slag*age + fly_ash*water + 
                  fly_ash*superplasticizer + 
                  fly_ash*coarse_agg + fly_ash*fine_agg + fly_ash*age + 
                  water*superplasticizer + water*coarse_agg + water*fine_agg+
                  water*age + superplasticizer*coarse_agg + 
                  superplasticizer*fine_agg+
                  superplasticizer*age + coarse_agg*fine_agg + coarse_agg*age
                  +fine_agg*age + I(cement^2) + I(blast_furnace_slag^2) + 
                  I(fly_ash^2) + I(water^2) + I(superplasticizer^2) + 
                  I(coarse_agg^2) + I(fine_agg^2) + I(age^2) + I(cement^3) + 
                  I(blast_furnace_slag^3) + I(fly_ash^3) + I(water^3) + 
                  I(superplasticizer^3) + I(coarse_agg^3) + I(fine_agg^3) + 
                  I(age^3) + I(sqrt(cement)) + I(sqrt(blast_furnace_slag)) + 
                  I(sqrt(fly_ash)) + I(sqrt(water)) + 
                  I(sqrt(superplasticizer)) + I(sqrt(coarse_agg)) + 
                  I(sqrt(fine_agg)) + I(sqrt(age)) + I(log(cement+1)) + 
                  I(log(blast_furnace_slag+1)) + 
                  I(log(fly_ash+1)) + I(log(water+1)) + 
                  I(log(superplasticizer+1)) + I(log(coarse_agg)) + 
                  I(log(fine_agg+1)) + I(log(age)),   
                  direction = "forward", k = log(n), trace = 0)

summary(fit_forw_bic)

press_fit_forw_bic = sqrt(sum((resid(fit_forw_bic)/(1-hatvalues(fit_forw_bic)))^2)/n)
press_fit_forw_bic

mse_ffbic = calculate_mse_test(fit_forw_bic, test)
mse_ffbic
```



```{r}
fit_null = lm(concrete_strength~1,data=train)
fit_step_aic = step(fit_null, 
                  scope = concrete_strength ~ cement + blast_furnace_slag +
                  fly_ash 
                  + water + superplasticizer + coarse_agg + fine_agg + age +
                  cement*blast_furnace_slag + cement*fly_ash + cement*water +
                  cement*superplasticizer + cement*coarse_agg 
                  +cement*fine_agg +
                  cement*age + blast_furnace_slag*fly_ash + 
                  blast_furnace_slag*fly_ash+
                  blast_furnace_slag*water + 
                  blast_furnace_slag*superplasticizer + 
                  blast_furnace_slag*coarse_agg + blast_furnace_slag*fine_agg
                  +blast_furnace_slag*age + fly_ash*water + 
                  fly_ash*superplasticizer + 
                  fly_ash*coarse_agg + fly_ash*fine_agg + fly_ash*age + 
                  water*superplasticizer + water*coarse_agg + water*fine_agg+
                  water*age + superplasticizer*coarse_agg + 
                  superplasticizer*fine_agg+
                  superplasticizer*age + coarse_agg*fine_agg + coarse_agg*age
                  +fine_agg*age + I(cement^2) + I(blast_furnace_slag^2) + 
                  I(fly_ash^2) + I(water^2) + I(superplasticizer^2) + 
                  I(coarse_agg^2) + I(fine_agg^2) + I(age^2) + I(cement^3) + 
                  I(blast_furnace_slag^3) + I(fly_ash^3) + I(water^3) + 
                  I(superplasticizer^3) + I(coarse_agg^3) + I(fine_agg^3) + 
                  I(age^3) + I(sqrt(cement)) + I(sqrt(blast_furnace_slag)) + 
                  I(sqrt(fly_ash)) + I(sqrt(water)) + 
                  I(sqrt(superplasticizer)) + I(sqrt(coarse_agg)) + 
                  I(sqrt(fine_agg)) + I(sqrt(age)) + I(log(cement+1)) + 
                  I(log(blast_furnace_slag+1)) + 
                  I(log(fly_ash+1)) + I(log(water+1)) + 
                  I(log(superplasticizer+1)) + I(log(coarse_agg)) + 
                  I(log(fine_agg+1)) + I(log(age)),   
                  direction = "both", trace = 0)

summary(fit_step_aic)

press_fit_step_aic = sqrt(sum((resid(fit_step_aic)/(1-hatvalues(fit_step_aic)))^2)/n)
press_fit_step_aic

mse_fsaic = calculate_mse_test(fit_step_aic, test)
mse_fsaic
```



#### 
```{r}
fit_null = lm(concrete_strength~1,data=train)
fit_step_bic = step(fit_null, 
                  scope = concrete_strength ~ cement + blast_furnace_slag +
                  fly_ash 
                  + water + superplasticizer + coarse_agg + fine_agg + age +
                  cement*blast_furnace_slag + cement*fly_ash + cement*water +
                  cement*superplasticizer + cement*coarse_agg 
                  +cement*fine_agg +
                  cement*age + blast_furnace_slag*fly_ash + 
                  blast_furnace_slag*fly_ash+
                  blast_furnace_slag*water + 
                  blast_furnace_slag*superplasticizer + 
                  blast_furnace_slag*coarse_agg + blast_furnace_slag*fine_agg
                  +blast_furnace_slag*age + fly_ash*water + 
                  fly_ash*superplasticizer + 
                  fly_ash*coarse_agg + fly_ash*fine_agg + fly_ash*age + 
                  water*superplasticizer + water*coarse_agg + water*fine_agg+
                  water*age + superplasticizer*coarse_agg + 
                  superplasticizer*fine_agg+
                  superplasticizer*age + coarse_agg*fine_agg + coarse_agg*age
                  +fine_agg*age + I(cement^2) + I(blast_furnace_slag^2) + 
                  I(fly_ash^2) + I(water^2) + I(superplasticizer^2) + 
                  I(coarse_agg^2) + I(fine_agg^2) + I(age^2) + I(cement^3) + 
                  I(blast_furnace_slag^3) + I(fly_ash^3) + I(water^3) + 
                  I(superplasticizer^3) + I(coarse_agg^3) + I(fine_agg^3) + 
                  I(age^3) + I(sqrt(cement)) + I(sqrt(blast_furnace_slag)) + 
                  I(sqrt(fly_ash)) + I(sqrt(water)) + 
                  I(sqrt(superplasticizer)) + I(sqrt(coarse_agg)) + 
                  I(sqrt(fine_agg)) + I(sqrt(age)) + I(log(cement+1)) + 
                  I(log(blast_furnace_slag+1)) + 
                  I(log(fly_ash+1)) + I(log(water+1)) + 
                  I(log(superplasticizer+1)) + I(log(coarse_agg)) + 
                  I(log(fine_agg+1)) + I(log(age)),   
                  direction = "both", trace = 0)

summary(fit_step_bic)

press_fit_step_bic = sqrt(sum((resid(fit_step_bic)/(1-hatvalues(fit_step_bic)))^2)/n)
press_fit_step_bic

mse_fsbic = calculate_mse_test(fit_step_bic, test)
mse_fsbic
```


#### Plotting metrics for models after variable reduction
```{r}
tab2 <- matrix(c('M8', 'M9', 'M10', 'M11', 'M12', 'M13',
                'FitBack_AIC', 'FitBack_BIC', 'FitForward_AIC', 
                'FitForward_BIC', 
                'FitStep_AIC' , 'FitStep_BIC',
                fit_back_aic$rank-1, fit_back_bic$rank-1, 
                fit_forw_aic$rank-1, fit_forw_bic$rank-1, 
                fit_step_aic$rank-1,
                fit_step_bic$rank-1, 
                summary(fit_back_aic)$r.squared, 
                summary(fit_back_bic)$r.squared, 
                summary(fit_forw_aic)$r.squared, 
                summary(fit_forw_bic)$r.squared, 
                summary(fit_step_aic)$r.squared,
                summary(fit_step_bic)$r.squared, 
                summary(fit_back_aic)$adj.r.squared, 
                summary(fit_back_bic)$adj.r.squared, 
                summary(fit_forw_aic)$adj.r.squared, 
                summary(fit_forw_bic)$adj.r.squared, 
                summary(fit_step_aic)$adj.r.squared,
                summary(fit_step_bic)$adj.r.squared,
                mse_fbaic, mse_fbbic, mse_ffaic, mse_ffbic, mse_fsaic, 
                mse_fsbic,
                press_fit_back_aic, press_fit_back_bic, press_fit_forw_aic, 
                press_fit_forw_bic, press_fit_step_aic, press_fit_step_bic
                ), ncol=7)
colnames(tab2) <- c('Model_Name', 'Model_Description','No_of_predictors','R2', 'Adj_R2', 'MSE', 'PRESS')
tab2 <- as.table(tab2)

metrics_df2 = as.data.frame.matrix(tab2) 

metrics_df2$R2 = as.numeric(as.character(metrics_df2$R2))
metrics_df2$Adj_R2 = as.numeric(as.character(metrics_df2$Adj_R2))
metrics_df2$MSE = as.numeric(as.character(metrics_df2$MSE))
metrics_df2$PRESS = as.numeric(as.character(metrics_df2$PRESS))

metrics_df2 = metrics_df2 %>% mutate(across(is.numeric, round, digits=2))

metrics_df2$Model_Name <- factor(metrics_df2$Model_Name, levels = metrics_df2$Model_Name)

ggplot(metrics_df2, aes(x=Model_Name, y=as.numeric(R2))) + geom_line(aes(group=1)) + geom_point()

ggplot(metrics_df2, aes(x=Model_Name, y=as.numeric(Adj_R2))) + geom_line(aes(group=1)) + geom_point()

ggplot(metrics_df2, aes(x=Model_Name, y=as.numeric(MSE))) + geom_line(aes(group=1)) + geom_point()

ggplot(metrics_df2, aes(x=Model_Name, y=as.numeric(PRESS))) + geom_line(aes(group=1)) + geom_point()

```


### K Fold Cross Validation for the selected models

K-Fold cross validation is useful for understanding how well the model generalises on the data.\

For this experiment, we will select some of the most promising models from the previous experiments and compare their performance over 5 fold cross validation.\

Selected Models:\
1. M1: Baseline Model\
2. M7: Baseline+Squared+Cubic+SquareRoot+Log+Interaction Model\
3. M10: Fit Forward AIC\
4. M9: Fit Backward BIC\
5. M12: Fit Step AIC\
6. M13: Fit Step BIC\
```{r}
k=8

#m1 baseline M1
#m2 all M7
#m3 ff aic M10
#m4 fb bic M9
#m5 fs aic M12
#m6 fs bic M13

RMSE_m1 = RMSE_m2 = RMSE_m3 = RMSE_m4 = RMSE_m5 = RMSE_m6 = numeric(k)

#Create k equally size folds
folds <- cut(1:n,breaks=k,labels=FALSE)

#Perform a k-fold cross validation
for(i in 1:k)
{
  # Find the indices for test data
  test_index = which(folds==i)
  
  # Obtain training/test data
  test_data = data[test_index, ]
  training_data = data[-test_index, ]
  
  
  model_1 = lm(concrete_strength ~ cement + blast_furnace_slag + 
                 fly_ash + water + superplasticizer + coarse_agg + 
                 fine_agg + age, data=training_data)
  
  model_2 = lm(concrete_strength ~ cement + blast_furnace_slag + fly_ash + 
                 water + superplasticizer + coarse_agg + fine_agg + age +
                 cement:blast_furnace_slag + cement:fly_ash + cement:water +
                 cement:superplasticizer + cement:coarse_agg + 
                 cement:fine_agg +
                 cement:age + blast_furnace_slag:fly_ash + 
                 blast_furnace_slag:fly_ash+
                 blast_furnace_slag:water + 
                 blast_furnace_slag:superplasticizer + 
                 blast_furnace_slag:coarse_agg + blast_furnace_slag:fine_agg 
                 +blast_furnace_slag:age + fly_ash:water + 
                 fly_ash:superplasticizer + 
                 fly_ash:coarse_agg + fly_ash:fine_agg + fly_ash:age + 
                 water:superplasticizer + water:coarse_agg + water:fine_agg+
                 water:age + superplasticizer:coarse_agg + 
                 superplasticizer:fine_agg+
                 superplasticizer:age + coarse_agg:fine_agg + coarse_agg:age 
                 +fine_agg:age + I(cement^2) + I(blast_furnace_slag^2) + 
                 I(fly_ash^2) + I(water^2) + I(superplasticizer^2) + 
                 I(coarse_agg^2) + 
                 I(fine_agg^2) + I(age^2) + I(cement^3) + 
                 I(blast_furnace_slag^3) + 
                 I(fly_ash^3) + I(water^3) + I(superplasticizer^3) + 
                 I(coarse_agg^3) + 
                 I(fine_agg^3) + I(age^3) + I(sqrt(cement)) + 
                 I(sqrt(blast_furnace_slag)) + 
                 I(sqrt(fly_ash)) + I(sqrt(water)) + 
                 I(sqrt(superplasticizer)) + I(sqrt(coarse_agg)) + 
                 I(sqrt(fine_agg)) + I(sqrt(age)) + I(log(cement+1)) + 
                 I(log(blast_furnace_slag+1)) + 
                 I(log(fly_ash+1)) + I(log(water+1)) + 
                 I(log(superplasticizer+1)) + I(log(coarse_agg)) + 
                 I(log(fine_agg+1)) + I(log(age)), data=training_data)
  
  model_3 = lm(concrete_strength~ I(log(age))+ cement+ I(log(superplasticizer                  + 1))+ blast_furnace_slag+ water+ superplasticizer+ 
                 I(log(fly_ash + 1))+ I(age^2)+ I(log(cement + 1))+ 
                 I(superplasticizer^2)+ 
                 I(superplasticizer^3)+ I(cement^3)+ I(sqrt(cement))+ 
                 I(blast_furnace_slag^3)+ 
                 fly_ash+ I(log(fine_agg + 1))+ I(sqrt(fine_agg))+ 
                 I(log(coarse_agg))+ 
                 I(age^3)+ I(water^3)+ I(log(water + 1))+ I(sqrt(water))+ 
                 fine_agg+ I(sqrt(age))+ I(cement^2)+ I(sqrt(fly_ash))+ 
                 I(fly_ash^2)+ 
                 I(log(blast_furnace_slag + 1))+ I(sqrt(blast_furnace_slag))+
                 I(blast_furnace_slag^2)+ I(fine_agg^2)+ I(fine_agg^3)+ 
                 I(sqrt(coarse_agg))+ 
                 I(fly_ash^3) + water:superplasticizer + 
                 blast_furnace_slag:superplasticizer + 
                 cement:water + blast_furnace_slag:fly_ash + water:fly_ash + 
                 cement:fine_agg +
                 cement:blast_furnace_slag, data=training_data)
  
  model_4 = lm(concrete_strength~ cement+ blast_furnace_slag+ fly_ash+ 
               water+ superplasticizer+ coarse_agg+ fine_agg+ age+ 
               I(blast_furnace_slag^2)+ 
               I(fly_ash^2)+ I(fine_agg^2)+ I(cement^3)+ 
               I(blast_furnace_slag^3)+ 
               I(fine_agg^3)+ I(sqrt(cement))+ I(sqrt(blast_furnace_slag))+ 
               I(sqrt(fly_ash))+ I(sqrt(water))+ I(sqrt(fine_agg))+ 
               I(log(cement + 1))+ I(log(blast_furnace_slag + 1))+ 
               I(log(fly_ash + 1))+ I(log(superplasticizer + 1))+ I(log(age))                +cement:water + cement:fine_agg + blast_furnace_slag:water + 
               blast_furnace_slag:superplasticizer + blast_furnace_slag:age +                fly_ash:water + fly_ash:superplasticizer + fly_ash:age + 
               water:superplasticizer + water:coarse_agg + water:fine_agg,                  data=training_data)
  
  model_5 = lm(concrete_strength~ I(log(age))+ cement+ blast_furnace_slag+ 
               water+ superplasticizer+ I(age^2)+ I(superplasticizer^2)+ 
               I(superplasticizer^3)+ I(cement^3)+ I(sqrt(cement))+ 
               I(blast_furnace_slag^3)+ 
               fly_ash+ I(log(fine_agg + 1))+ I(sqrt(fine_agg))+ 
               I(log(coarse_agg))+ 
               I(age^3)+ I(water^3)+ I(log(water + 1))+ I(sqrt(water))+ 
               fine_agg+ I(sqrt(age))+ I(cement^2) + water:superplasticizer +
               blast_furnace_slag:superplasticizer + cement:water + 
               water:fly_ash +
               cement:fine_agg + water:fine_agg, data=training_data)
  
  model_6 = lm(concrete_strength~ I(log(age))+ cement+ blast_furnace_slag+ 
               water+ superplasticizer+ I(log(fly_ash + 1))+ I(age^2)+ 
               I(log(cement + 1))+ I(superplasticizer^2)+ 
               I(superplasticizer^3) + 
               water:superplasticizer + blast_furnace_slag:superplasticizer +
               cement:water, data=training_data)
  
  # Obtain RMSE on the 'test' data
  resid_m1 = test_data["concrete_strength"] - predict(model_1, newdata=test_data)              
  RMSE_m1[i] = sqrt(sum(resid_m1^2)/nrow(test_data)) 
  
  resid_m2 = test_data[,"concrete_strength"] - predict(model_2, newdata=test_data) 
  RMSE_m2[i] = sqrt(sum(resid_m2^2)/nrow(test_data))
  
  resid_m3 = test_data[,"concrete_strength"] - predict(model_3, newdata=test_data) 
  RMSE_m3[i] = sqrt(sum(resid_m3^2)/nrow(test_data))
  
  resid_m4 = test_data[,"concrete_strength"] - predict(model_4, newdata=test_data) 
  RMSE_m4[i] = sqrt(sum(resid_m4^2)/nrow(test_data))
  
  resid_m5 = test_data[,"concrete_strength"] - predict(model_5, newdata=test_data) 
  RMSE_m5[i] = sqrt(sum(resid_m5^2)/nrow(test_data))
  
  resid_m6 = test_data[,"concrete_strength"] - predict(model_6, newdata=test_data) 
  RMSE_m6[i] = sqrt(sum(resid_m6^2)/nrow(test_data))
}


cat("Baseline model: ", mean(RMSE_m1))
cat("Complete  model with both all interaction and polynomial terms: ", mean(RMSE_m2))
cat("Fit Forward AIC model: ",mean(RMSE_m3))
cat("Fit Backward BIC model: ",mean(RMSE_m4))
cat("Fit Stepwise AIC model: ",mean(RMSE_m5))
cat("Fit Stepwise BIC model: ",mean(RMSE_m6))

```

From all these experiments, we can conclude that the model we obtain after applying Stepwise AIC on the complete model(polynomial + interaction) is the best model for predicting the strength of concrete.


## Model Interpretability

### Variable Importance

#### Using Standardized Model Coefficients
```{r}
data_std = as.data.frame(scale(data, center=TRUE, scale=TRUE))

mlr_std <- lm(concrete_strength~.,data=data_std)
summary(mlr_std)

sort(abs(mlr_std$coefficients), decreasing =TRUE)
```

Top influential features : cement, blast_furnace, age, fly_ash


#### Random Forest Variable Importance
```{r}

rf <- randomForest(concrete_strength~., data=data, proximity=TRUE)
summary(rf)
importance(rf)
varImpPlot(rf)
```

Top influential features : age, cement, water, superplasticizer

#### R2 from Single Predictor Model
```{r}
calculate_r2 <- function(var_name, data) {
  print(var_name)
  fm <- as.formula(paste("concrete_strength", "~", var_name))
  model = lm(fm, data = data)
  r2 = summary(model)$r.squared
  return(r2)
}


for(el in names(data)) {
  if (el != "concrete_strength")
  {
  r2 = calculate_r2(el, data)
  cat("R2 is:  ", r2, "\n")
  }
}
```

Top influential features : cement, superplasticizer, age, water

Based on all these experiments, we can conclude that the following are highly likely to be features of significant importance to the model:

1. Cement
2. Superplasticizer
3. Water
4. Age


### Final Balanced Model

We choose the Model we got by using Stepwise BIC as a balanced model. It has 13 predictors as well as considerably high performance metrics.
```{r}
best_model = lm(concrete_strength~ I(log(age))+ cement+ blast_furnace_slag+ 
               water+ superplasticizer+ I(log(fly_ash + 1))+ I(age^2)+ 
               I(log(cement + 1))+ I(superplasticizer^2)+ 
               I(superplasticizer^3) + 
               water:superplasticizer + blast_furnace_slag:superplasticizer +
               cement:water, data=train)
summary(best_model)

vif(best_model)
```

Dropping interaction and polynomial terms with high VIF\
\
Features Dropped:\
- cement:water\
- blast_furnace_slag:superplasticizer\
- water:superplasticizer\
- superplasticizer^2\
- log(cement)\

```{r}
best_model = lm(concrete_strength~ I(log(age))+ cement+ blast_furnace_slag+ 
               water+ superplasticizer+ I(log(fly_ash + 1))+ I(age^2)+ 
               I(superplasticizer^3), data=train)
summary(best_model)

vif(best_model)

press_best = sqrt(sum((resid(best_model)/(1-hatvalues(best_model)))^2)/n)
press_best

mse_best = calculate_mse_test(best_model, test)
mse_best
```
This model gives us a good balance of predictability as well as interpretability.


### Checking Model Assumptions for the Final Selected model
```{r}
check_model_assumptions(best_model)
```
The normality assumption holds true, as the p-value for the Shapiro Wilks test is greater than the significance level (0.05).

The equal variance assumption fails as the p value for the BP test is still less than 0.05.

The linearity assumption seems to get violated as the data points are not equally distributed on either side of the axis. 



### Significance of Regression (Testing Hypothesis)

#### Verifing initial assumptions

We can use the final model to verify whether some of the initial hypothesis we formulated hold true.

1. Water is a significant predictor for the model
2. Cement is a significant variable for predicting concrete strength.
3. Fine aggregate and coarse aggregate do not contribute significantly   towards the prediction of concrete strength.
```{r}
summary(best_model)

summary(baseline)

```

By looking at the model summary for both the best model and the baseline model, we can conclude that:

1. Water is an important predictor. [p-value < 0.05]
2. Cement is an important predictor. [p-value < 0.05]
3. Fine aggregate and coarse aggregate are not significant predictors. [p-value > 0.05 and not in best model]


  



